Internal physiology and external cues drive changes in reproductive axis of parenting pigeons

### 2020-03-06 analysis/figure-5.Rmd
save files
now with manuscritp draft

allDEG <- DEG_pathfiles %>%
  setNames(nm = .) %>% 
  map_df(~read_csv(.x), .id = "file_name") %>% 
  mutate(DEG = sapply(strsplit(as.character(file_name),'./results/DEseq2/manip/'), "[", 2))  %>% 
  mutate(DEG = sapply(strsplit(as.character(DEG),'_diffexp.csv'), "[", 1))  %>% 
  mutate(tissue = sapply(strsplit(as.character(DEG),'\\.'), "[", 1)) %>%
  mutate(down = sapply(strsplit(as.character(DEG),'\\_'), "[", 3)) %>%
  mutate(up = sapply(strsplit(as.character(DEG),'\\_'), "[", 4)) %>%
  mutate(comparison = paste(down,up, sep = "_")) %>%
  mutate(sex = sapply(strsplit(as.character(sextissue),'\\_'), "[", 1)) %>%
  mutate(tissue = sapply(strsplit(as.character(sextissue),'\\_'), "[", 2)) %>%
dplyr::select(sex,tissue,comparison, direction, gene, lfc, padj, logpadj) 
head(allDEG)

comparisonlevels <- c( "inc.d9_m.inc.d8", "inc.d17_prolong", "hatch_extend" , 
                       "inc.d3_m.inc.d3" ,  "inc.d9_m.inc.d9", 
                        "inc.d17_m.inc.d17", "m.n2_hatch" 
                        )

@@

removalplots <- allDEG %>%
  filter(comparison %in% c("inc.d3_m.inc.d3" ,  "inc.d9_m.inc.d9", 
                        "inc.d17_m.inc.d17", "m.n2_hatch" ))  %>%
ggplot(aes(x = comparison,  fill = direction)) +
  geom_bar(position = "dodge") +
  facet_grid(tissue~sex) +
  theme_B3() +
  theme(#axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text.x = element_blank(),
         axis.text.y = element_blank(),
        legend.position = "none",
        strip.placement = "right")  +
  guides(fill = guide_legend(nrow = 1)) +
  labs(x = "Offspring removal", y = NULL) +
  scale_fill_manual(values = allcolors,
                       name = "increased in", drop = FALSE) +
  scale_color_manual(values = sexcolors) +
  geom_text(stat='count', aes(label=..count..), vjust =-0.5, 
            position = position_dodge(width = 1),
            size = 2.5, color = "black") +
  ylim(0,3500)

@@

fig5 <- plot_grid(expdesign, barplots, mylegend, ncol =1, rel_heights = c(0.25,1,0.1))
fig5
```

>>>>

allDEG <- DEG_pathfiles %>%
  setNames(nm = .) %>% 
  map_df(~read_csv(.x), .id = "file_name") %>% 
  mutate(DEG = sapply(strsplit(as.character(file_name),'./results/DEseq2/manip/'), "[", 2))  %>% 
  mutate(DEG = sapply(strsplit(as.character(DEG),'_diffexp.csv'), "[", 1))  %>% 
  mutate(tissue = sapply(strsplit(as.character(DEG),'\\.'), "[", 1)) %>%
  mutate(down = sapply(strsplit(as.character(DEG),'\\_'), "[", 3)) %>%
  mutate(up = sapply(strsplit(as.character(DEG),'\\_'), "[", 4)) %>%
  mutate(comparison = paste(down,up, sep = "_")) %>%
  mutate(sex = sapply(strsplit(as.character(sextissue),'\\_'), "[", 1)) %>%
  mutate(tissue = sapply(strsplit(as.character(sextissue),'\\_'), "[", 2))

allDEG <- allDEG %>%
dplyr::select(sex,tissue,comparison, direction, gene, lfc, padj, logpadj)

allDEG %>% distinct(comparison)

comparisonlevels <-  c(	 "inc.d3_m.inc.d3",			
                             "inc.d9_m.inc.d9",	
                             "inc.d17_m.inc.d17",		
                           "hatch_m.n2",

                           "inc.d9_m.inc.d8",

                         "m.inc.d9_m.n2",
                         "m.inc.d9_m.inc.d8",

                         "m.inc.d8_extend",
                         "prolong_extend",
                           "hatch_extend",
                         "m.n2_extend",

                         "m.inc.d8_prolong",
                            "inc.d17_prolong",	
                           "m.inc.d17_prolong",

                           "m.inc.d3_m.inc.d9",
                           "m.inc.d3_m.inc.d17",
                         "m.inc.d3_m.n2",
                         "m.inc.d9_m.inc.d17")

@@

removalplots <- allDEG %>%
  filter(comparison %in% c("inc.d3_m.inc.d3" ,  "inc.d9_m.inc.d9", 
                        "inc.d17_m.inc.d17", "hatch_m.n2" ))  %>%
ggplot(aes(x = comparison,  fill = direction)) +
  geom_bar(position = "dodge") +
  facet_grid(tissue~sex) +
  theme_B3() +
  theme(#axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text.x = element_blank(),
         axis.text.y = element_blank(),
        legend.position = "none",
        strip.placement = "right")  +
  guides(fill = guide_legend(nrow = 1)) +
  labs(x = "Offspring removal", y = NULL) +
  scale_fill_manual(values = allcolors,
                       name = "increased in", drop = FALSE) +
  scale_color_manual(values = sexcolors) +
  geom_text(stat='count', aes(label=..count..), vjust =-0.5, 
            position = position_dodge(width = 1),
            size = 2.5, color = "black") +
  ylim(0,3500)

@@

fig5 <- plot_grid(expdesign, barplots, ncol =1, rel_heights = c(0.5,1))
fig5

```

```{r writefiles}
write_csv(allDEG, "../results/DEGs-manip.csv")
```

## subset. examine for E, T, hyp, and gonad hypotheses

```{r subset}

subsetDEGs <- function(whichsex, whichtissue, whichcomparison) {
df <- allDEG %>%
  filter(sex %in% whichsex,
         tissue %in% whichtissue,
         comparison %in% whichcomparison)  %>%
  group_by(sex, tissue, comparison) #%>%
  #summarize(DEGs = n()) %>%
  #arrange(desc(DEGs)) %>%
  #drop_na()
print(df)
}

subsetDEGs(c("male","female"), c("hypothalamus", "pituitary", "gonad"), comparisonlevels)
fhyphatchext <- subsetDEGs("female", "hypothalamus", "hatch_extend") %>% arrange (gene)
mhyphatchext <- subsetDEGs("male", "hypothalamus", "hatch_extend") %>% arrange (gene)

ggplot(fhyphatchext, aes(x = lfc, y = logpadj)) + geom_point() + xlim(-5,5) + ylim(0,7)
ggplot(mhyphatchext, aes(x = lfc, y = logpadj)) + geom_point() + xlim(-5,5) + ylim(0,7)

```

### 2020-03-06 docs/manuscript/elife_v1.Rmd
now with manuscritp draft

% !TeX root = RJwrapper.tex
\title{Internal physiology and external cues drive changes in reproductive axis
of parenting pigeons}
\author{by Rayna M Harris \footnote{equal contribtion}, Suzanne H. Austin \footnote{equal contribtion}, Andrew S. Lang, Victoria S. Farrar, April Booth, Tanner Feustel, Matthew D. MacManes, Rebecca M. Calisi \footnote{corresponding author}}

\maketitle

% Any extra LaTeX you need in the preamble

In species that provide parental care, successful rearing of offspring involves a shift from aggressive and sexual behaviors to more caring and nurturing ones.  We hypothesize that these shifts in behavior are related to changes in gene expression that are controlled by internal mechanisms rather than by the external cues.  We found that gene expression in the hypothalamus is fairly insensitive to the natural transition between egg and nestling stages; however, in the pituitary and gonads, gene expression showed much more plasticity over the course of reproduction. We show evidence that natural transitions in reproductive and parental behavior by mediated internal clock controlling gene expression in the pitary and gonads but that the hypothalamus is very sensitive to external cues that signal a distruption in the natural parental care cycle. This research provides a deeper understanding of the interplay of genes, hormones, and parental care and has important implications for understanding molecular processes that underlie behavior transitions.

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

Reproduction is a critical period in an organism's life cycle. Parents
must balance their own needs with the energetic costs and
time-constraints required to successfully reproduce and rear their
offspring to independence(1). A complex series of evolutionarily
conserved physiological processes underlie behaviors typically
associated with reproduction in vertebrates, including courtship,
nesting, and parental care. These processes are primarily mediated by
the reproductive axis, which consists of the hypothalamus in the brain,
the pituitary, and the gonads (i.e., the HPG). Our understanding of the
physiological changes that adults navigate as they transition through
reproductive states are primarily limited to a number of circulating
hormones and a small set of well-known genes (cite). We have yet to come
close to conceptualizing the larger symphony of reproductive events that
occur to yield reproductive and parental care behaviors.

Using the model of a biparental vertebrate species, the rock dove
(Columba livia), our team has previously demonstrated the dynamic nature
of gene activity within the reproductive or HPG axis.
\citet{MacManes2017} reported that HPG gene activity varied by sex, with
females exhibiting higher levels of overall gene expression than males.
\citet{Calisi2018} reported that HPG gene activity varied response to
restraint-stress, and Austin et al.~identify genes that responds to both
corticosterone treatment and restraint stress. While each of these
studies advanced our understanding of sex-differences and
transcriptional plasticity in the reproducitve axis, none of these
studies examined activity of the reproducitve axis of bird that are
actively reproducing or caring for offspring. Studying parental care in
both sexes of mammalian species is challenging because only the females
are able to use lactation to feed their young, and the physiological
mechanisms regulating this process are hard to control for. The rock
dove is an excellent system for studying reproduction and parental care
in both sexes because both males and females feed their young an
energy-rich secretion produced from the epithelial lining of their
crops, known as crop milk.

@@

\hypertarget{results}{%
\subsection{Results}\label{results}}

@@

Our results suggest that tissue is the greatest source of variation in
the dataset, with hypothalamic, pituitary, and gonadal samples all
forming distinct clusters, with only a few pituitary and gonad samples
falling outside the 95\% confidence interval (Fig. \ref{figure:fig1}b).
Sex-differences are most prounounced in the gonads with male and female
samples form two discrete clusters (Fig. \ref{figure:fig1}c). Treatment
accounts for much less of the variation than tissue and sex, and even
after removing sex as a variatble, treatment effects in all three
tissues are hard to visualize simultatenously (Fig.
\ref{figure:fig1}d,e), so we next looked at each tissue and sex
separately. In the hypothalamus, control samples form tight clusters
that stands out from all other timepoints that are largely overlapping
in space (Fig. \ref{figure:fig1}h). The pituitary shows the strongest
treatment effect with one cluster of the ``earlier'' parental timepoints
(nest-building, lay, incubation days 3 and 9), a second cluster of
``later'' parental timpoints (incubation day 17, hatch, and nestling
care days 5 and 9; Fig. \ref{figure:fig1}i). Like the hypothalamus, in
the gonads, the control timepoints appear to be much less variable that
the parental timepoints, but there is no clear separation of gonadal
samples by parental stage (Fig. \ref{figure:fig1}j). While the tSNE
analysis reveals broad patterns across all levels of the experiment,
additional tools are needed it hone in on the specific differences
associated with parental care in each tissue.

@@

\hypertarget{discussion}{%
\subsection{Discussion}\label{discussion}}

\hypertarget{materials-and-methods}{%
\subsection{Materials and Methods}\label{materials-and-methods}}

\hypertarget{rnaseq}{%
\subsubsection{RNAseq}\label{rnaseq}}

@@

\hypertarget{summary}{%
\subsection{Summary}\label{summary}}

This file is only a basic article template. For full details of
\emph{The R Journal} style and information on how to prepare your
article for submission, see the
\href{https://journal.r-project.org/share/author-guide.pdf}{Instructions
for Authors}.

\hypertarget{acknowledgements}{%
\subsection{Acknowledgements}\label{acknowledgements}}

@@

\hypertarget{references}{%
\subsection{References}\label{references}}

@@

\address{%
Rayna M Harris \footnote{equal contribtion}\\
UC Davis\\
\\
}

\address{%
Suzanne H. Austin \footnote{equal contribtion}\\
UC Davis\\
\\
}

\address{%
Andrew S. Lang\\
UNH\\
\\
}

\address{%
Victoria S. Farrar\\
UC Davis\\
\\
}

\address{%
April Booth\\
UC Davis\\
\\
}

\address{%
Tanner Feustel\\
UC Davis\\
\\
}

\address{%
Matthew D. MacManes\\
UNH\\
\\
}

\address{%
Rebecca M. Calisi \footnote{corresponding author}\\
UC Davis\\
\\
}
\href{mailto:author7@work}{\nolinkurl{author7@work}}

>>>>

---
title: Internal physiology and external cues drive changes in reproductive axis of parenting pigeons
author:
  - name: Rayna M Harris ^[equal contribtion] 
    affiliation: UC Davis
  - name: Suzanne H. Austin  ^[equal contribtion] 
    affiliation: UC Davis
  - name: Andrew S. Lang
    affiliation: UNH
  - name: Victoria S. Farrar
    affiliation: UC Davis
  - name: April Booth
    affiliation: UC Davis
  - name: Tanner Feustel
    affiliation: UC Davis
  - name: Matthew D. MacManes
    affiliation: UNH
  - name: Rebecca M. Calisi  ^[corresponding author] 
    affiliation: UC Davis
    email:  author7@work   
address:
  - code: UC Davis
    address: Department, Street, City, CA, Zip
  - code: UNH
    address: Department, Street, City, NH, Zip    
abstract: |
output:
  rticles::rjournal_article:
    includes:
      in_header: preamble.tex

---

## Introduction

Reproduction is a critical period in an organism’s life cycle. Parents must balance their own needs with the energetic costs and time-constraints required to successfully reproduce and rear their offspring to independence(1).  A complex series of evolutionarily conserved physiological processes underlie behaviors typically associated with reproduction in vertebrates, including courtship, nesting, and parental care. These processes are primarily mediated by the reproductive axis, which consists of the hypothalamus in the brain, the pituitary, and the gonads (i.e., the HPG). Our understanding of the physiological changes that adults navigate as they transition through reproductive states are primarily limited to a number of circulating hormones and a small set of well-known genes (cite). We have yet to come close to conceptualizing the larger symphony of reproductive events that occur to yield reproductive and parental care behaviors.

Using the model of a biparental vertebrate species, the rock dove (Columba livia), our team has previously demonstrated the dynamic nature of gene activity within the reproductive or HPG axis. \citet{MacManes2017} reported that HPG gene activity varied by sex, with females exhibiting higher levels of overall gene expression than males. \citet{Calisi2018} reported that HPG gene activity varied response to restraint-stress, and Austin et al. identify genes that responds to both corticosterone treatment and restraint stress. While each of these studies advanced our understanding of sex-differences and transcriptional plasticity in the reproducitve axis, none of these studies examined activity of the reproducitve axis of bird that are actively reproducing or caring for offspring. Studying parental care in both sexes of mammalian species is challenging because only the females are able to use lactation to feed their young, and the physiological mechanisms regulating this process are hard to control for. The rock dove is an excellent system for studying reproduction and parental care in both sexes because both males and females feed their young an energy-rich secretion produced from the epithelial lining of their crops, known as crop milk.

@@

## Results

@@

Our results suggest that tissue is the greatest source of variation in the dataset, with hypothalamic, pituitary, and gonadal samples all forming distinct clusters, with only a few pituitary and gonad samples falling outside the 95% confidence interval (Fig. \ref{figure:fig1}b). Sex-differences are most prounounced in the gonads with male and female samples form two discrete clusters (Fig. \ref{figure:fig1}c). Treatment accounts for much less of the variation than tissue and sex, and even after removing sex as a variatble, treatment effects in all three tissues are hard to visualize simultatenously (Fig. \ref{figure:fig1}d,e), so we next looked at each tissue and sex separately. In the hypothalamus, control samples form tight clusters that stands out from all other timepoints that are largely overlapping in space (Fig. \ref{figure:fig1}h). The pituitary shows the strongest treatment effect with one cluster of the “earlier” parental timepoints (nest-building, lay, incubation days 3 and 9), a second cluster of “later” parental timpoints (incubation day 17, hatch, and nestling care days 5 and 9; Fig. \ref{figure:fig1}i). Like the hypothalamus, in the gonads, the control timepoints appear to be much less variable that the parental timepoints, but there is no clear separation of gonadal samples by parental stage (Fig. \ref{figure:fig1}j). While the tSNE analysis reveals broad patterns across all levels of the experiment, additional tools are needed it hone in on the specific differences associated with parental care in each tissue.

@@

## Discussion

## Materials and Methods

### RNAseq

@@

## Summary

This file is only a basic article template. For full details of _The R Journal_ style and information on how to prepare your article for submission, see the [Instructions for Authors](https://journal.r-project.org/share/author-guide.pdf).

## Acknowledgements

@@

## References

@@

<!-- 
Introductory section which may include references in parentheses
\citep{RDevelopmentCoreTeam2013}, or cite a reference such as \citet{RDevelopmentCoreTeam2013} in the text.
-->
